---
title: "ggplot"
author: "Naomi Tague"
date: "Feb 2, 2016"
output: html_document
---

```{r setup, echo=FALSE}
#read in data
thin = read.table("thinning.csv", sep=",", header=T)
clim=read.csv("clim.csv")
#make sure you have libraries
library(dplyr)
library(tidyr)
library(ggplot2)
library(gridExtra)
library(lubridate)

```
*GGPLOT* is a "grammer of graphics", in other words a set of elements (like words, phrases) that you can put together to make a plot. A "grammer" is more flexible than standard plotting; but a bit harder to learn
In ggplot, we combine the following element:

* *data* a data set used to generate the plot

* *geoms* visual character of the plot- boxplot, scatter plot

* *aesthetics* what gets put into the geom

* *themes* characteristics of plotting environment (axis label font...)

*GGPLOT* is organized in a way that you can build a very complex plot by adding pieces to it

 A simple example

```{r basic example}
# basic plot
ggplot(clim, aes(y=tmax, x=month))+geom_point()
# save the basic plot so we can add to it
p = ggplot(clim, aes(y=tmax, x=month))+geom_point()
p
p+labs(y="Maximum Temperature", x="Month")
p = p + labs(y="Maximum Temperature", x="Month")
# themes are used to control character of the plot
pclr = p+theme(axis.text= element_text(face="bold", colour="red", size=14))

# to display more than one plot (a "matrix" or grid of plots)

grid.arrange(p,pclr)
# or to control to make by col
grid.arrange(p, pclr, ncol=2)

# there are also built in themes
pclr = pclr+theme_bw()
pclr
```
There are many different geoms (geometries) - lets try the standard ones

```{r plot types}

# note the use of as factor (boxplot need factors to organize)
p1=ggplot(clim, aes(y=tmax, x=as.factor(month)))+geom_boxplot(colour="green", fill="purple")
p1=p1+labs(y="Max Temp C", x="Month")
p1

# scatter plot
p2=ggplot(clim, aes(x=tmax, y=tmin))+geom_point(col="blue", shape=9, size=rel(4))
p2
  p2=p2+labs(x="Max Temp C", y="Min Temp C") 
  p2 = p2+ ggtitle("How does daily maximim and min temp compare")
  p2 = p2 + geom_abline(intercept=0,slope=1, colour="yellow", size=4)
p2

#density plot
p3=ggplot(clim, aes(x=rain))+geom_density()
p3
p3 = ggplot(subset(clim, clim$rain > 0), aes(x=rain))+geom_density(fill="blue")
#p4=ggplot(clim, aes(x=date,y=rain))+geom_line()+ggtitle("Line Graph")
p3
#fix issue with date
clim$date = mdy(paste(clim$month, clim$day, clim$year, sep="/"))
p4=ggplot(clim, aes(x=date,y=rain))+geom_line()+ggtitle("Line Graph")
p4
grid.arrange(p1,p2,p3,p4)
```
One of the most useful things about GGPLOT is that it makes it easy to visualize your data in ways that highlight different attributes - this can help you to see multiple dimensions at once
There are multiple ways to do this

* color/size/shape (within plots)
* facets  (between plots)

Lets start with color
We can use a more interesting data set *thin*



```{r color}
p1=ggplot(thin, aes(x=as.factor(postyears), y=cpool, col=as.factor(thin)))+geom_boxplot()
p1 = p1+labs(x="Years since thinning", y="Carbon Sequestration")
p1
p1 = p1 + scale_fill_discrete(name="Thinning %")

```

Not so pretty, but what if we look at mean over all possible startyears instead - averaging over climate variability
```{r aggregate}

p1 = ggplot(thin, aes(x=as.factor(postyears), y=cpool, fill=as.factor(postyears))) + stat_summary(fun.y="mean", geom="bar") + labs(x="Time since thinning", y="Carbon Sequestration")
p1 = p1 + theme(legend.position="none")
p1
p1 = ggplot(thin, aes(x=as.factor(postyears), y=cpool, col=as.factor(thin))) + stat_summary(fun.y="mean", geom="point")
p1
p1 = ggplot(thin, aes(x=postyears, y=cpool, col=as.factor(thin))) + stat_summary(fun.y="mean", geom="line")

```

Now lets try *facets* or different graphs and play a bit with aggregating data by summing
```{r facet}
p2 = ggplot(thin, aes(x=postyears, y=cpool))+stat_summary(fun.y="mean", geom="line")
p2= p2+facet_wrap(~thin)

p2 = ggplot(thin, aes(x=as.factor(postyears), y=cpool))+geom_boxplot()
p2
p2= p2+facet_wrap(~thin)
p2

p3 = ggplot(clim, aes(x=year,y=rain))+stat_summary(fun.y="sum", geom="bar", col="blue")
p3
p3 = ggplot(clim, aes(x=month,y=rain))+stat_summary(fun.y="mean", geom="bar")
p3
p3 + facet_wrap(~year)
```

