---
title: "joining"
author: "Ben Best"
date: "January 29, 2016"
output: html_document
---


### `bind_cols`

```{r bind_cols, eval=F}
y = data.frame(
  x1 = c('A','B','C'), 
  x2 = c( 1 , 2 , 3), 
  stringsAsFactors=F)
z = data.frame(
  x1 = c('B','C','D'), 
  x2 = c( 2 , 3 , 4), 
  stringsAsFactors=F)
y
z
bind_cols(y, z)
```

### `bind_rows`

```{r bind_rows, eval=F}
y
z
bind_rows(y, z)
```

### `union`

```{r union, eval=F}
y
z
union(y, z)
```

### `intersect`

```{r intersect, eval=F}
y
z
intersect(y, z)
```

### `setdiff`

```{r setdiff, eval=F}
y
z
setdiff(y, z)
```

### `left_join`

```{r left_join, eval=F}
songs = data.frame(
  song = c('Across the Universe','Come Together', 'Hello, Goodbye', 'Peggy Sue'),
  name = c('John','John','Paul','Buddy'), 
  stringsAsFactors=F)
artists = data.frame(
  name = c('George','John','Paul','Ringo'),
  plays = c('sitar','guitar','bass','drums'), 
  stringsAsFactors=F)
left_join(songs, artists, by='name')
```

### `inner_join`

```{r inner_join, eval=F}
inner_join(songs, artists, by = "name")
```

### `semi_join`

```{r semi_join, eval=F}
semi_join(songs, artists, by = "name")
```

### `anti_join`

```{r anti_join, eval=F}
anti_join(songs, artists, by = "name")
```

### summarize per capita CO<sub>2</sub> emissions 

You'll join the [gapminder](https://github.com/jennybc/gapminder) datasets to get world population per country.

_**Task**. Report the top 5 emitting countries (not World or EU28) per capita using your long format table. (You may need to convert your year column from factor to numeric, eg `mutate(year = as.numeric(as.character(year)))`. As with most analyses, there are multiple ways to do this. I used the following functions: `filter`, `arrange`, `desc`, `head`)_. 

_**Task**. Summarize the total emissions by country  (not World or EU28) per capita across years from your long format table and return the top 5 emitting countries. (As with most analyses, there are multiple ways to do this. I used the following functions: `filter`, `arrange`, `desc`, `head`)_. 

```{r gapminder, eval=F}
library(gapminder) # install.packages('gapminder')
```

