---
title: 'Week 4: Wrangling Data II'
author: "Ben Best"
date: '`r format(Sys.time(), "%Y-%m-%d %H:%M")`'
output: html_document
---

## Schedule

- 8:30 - 9:15 [BB] **demo** functions from dplyr, tidyr, lubridate, and stringr libraries by walking through examples all together

- 9:15 - 9:45 [BB] **individual** exercises answering questions from datasets on [CO2 emissions by country 1970-2014](http://edgar.jrc.ec.europa.eu/overview.php?v=CO2ts1990-2014&sort=des9)

- 9:45 - 10:00 **break**

- 10:00 - 10:30 [BB] **group** work asking questions and answering them with your own datasets

- 10:30 - 11:30 [NT] **programming concepts**

## Precursors

### Invitations

- **organizations**: invite @bbest and @ctague to your https://github.com/<organization> so we can directly _push_ to your site (vs _fork_ & _pull request_)
- **auditors**: add to GauchoSpace to use announcements there

## Materials

- [stat545-ubc: Data wrangling, exploration, and analysis with R](https://stat545-ubc.github.io/)

## Demo

nycflights13

```{r dplyr I, eval=F}
library(dplyr)
library(nycflights13)

# Convert data to a tbl_df so that it uses dplyr's nice print method
flights <- tbl_df(flights)
flights

# Use glimpse to few some values in each column
glimpse(flights)

# filter: inspect subsets of data
# How many flights flew to Madison in 2013?
flights %>%
  filter(dest == "MSN")

# filter: multiple conditions
# How many flights flew to Madison in first week of January?
# Comma separated conditions are combined with '&'
flights %>%
  filter(dest == "MSN", month == 1, day <= 7)

# QuestionCommas in the filter statement are implicit & (and) operators. Is there anything similar for | (or)?
# Logical or statements are supported, but there's no shorthand.
flights %>%
  filter(dest == "MSN" | dest == "ORD" | dest == "MDW")
# For more complicated checks, I would try a set operation.
flights %>%
  filter(dest %in% c("MSN", "ORD", "MDW"))


# arrange: sort columns
# Sort by which airport they departed from in NYC, then year, month, day.
flights %>%
  arrange(origin, year, month, day)

# desc: reverses sorting of a column
# Find longest delayed flights to Madison.
flights %>%
  filter(dest == "MSN") %>%
  arrange(desc(dep_delay))

# select
# Select the columns you want.
flights %>%
  select(origin, year, month, day)

# select's helpers
# select has many helper functions. See ?select.
flights %>%
  select(origin, year:day, starts_with("dep"))

# negative selecting
# We can drop columns by “negating” the name. Since helpers give us column names, we can negate them too.
flights %>%
  select(-dest, -starts_with("arr"),
         -ends_with("time"))
```

Recap: Verbs for inspecting data:

- convert to a `tbl_df` - nice print method
- `glimpse` - some of each column
- `filter` - subsetting
- `arrange` - sorting (desc to reverse the sort)
- `select` - picking (and omiting) columns

```{r dplyr II, eval=F}
# rename
# Rename columns with rename(NewName = OldName). To keep the order correct, read/remember the renaming = as “was”.
flights %>%
  rename(y = year, m = month, d = day)

# mutate
# How much departure delay did the flight make up for in the air?
flights %>%
  mutate(
    gain = arr_delay - dep_delay,
    speed = (distance / air_time) * 60,
    gain_per_hour = gain / (air_time / 60)) %>%
  select(gain:gain_per_hour)

# group_by
# Let's compute the average delay per month of flights to Madison.
# Normally–in aggregate, by or plyr's d*ply functions–you specify the grouping as an argument to the aggregation function.
aggregate(dep_delay ~ month, flights, mean,
          subset = flights$dest == "MSN")

# group_by
# In dplyr, grouping is its own action. It is done as its own step in the pipeline. Here, we filter to the flights to Madison and group_by month.
msn_by_month <- flights %>%
  filter(dest == "MSN") %>%
  group_by(month)
msn_by_month

# summarise
# Now we use summarise to compute (several) aggregate values within each group (month). summarise returns one row per group.
msn_by_month %>%
  summarise(
    flights = n(),
    avg_delay = mean(dep_delay, na.rm = TRUE),
    n_planes = n_distinct(tailnum))

# tally
# tally is a shortcut for counting number of items per group.
# Number of flights from NYC by destination by month:
flights %>%
  group_by(dest, month) %>%
  tally

# ungroup
# Remove the grouping structure with ungroup.
msn_by_month %>% ungroup

# Summarizing undoes grouping.
# Each summarise statement peels off one layer of grouping (from the right of the list of groups).
# day gets peeled off
per_day <- flights %>%
  group_by(dest, year, month, day) %>%
  summarise(flights = n())
per_day

# Peel off month grouping
per_month <- per_day %>%
  summarise(flights = sum(flights))
per_month
```


That covers 80% of dplyr:

- `select`
- `filter`
- `arrange`
- `glimpse`
- `rename`
- `mutate`
- `group_by`, `ungroup`
- `summarise`

Other 20%

- assembly: `bind_rows`, `bind_cols`
- column-wise operations: `mutate_each`, `summarise_each`
- join tables together: `left_join`, `right_join`, `inner_join`, `full_join`
- filtering joins: `semi_join`, `anti_join
- `do`: arbitrary code on each chunk
- different types of tabular data (databases, data.tables)

### tidyr

- what is an observation? wide vs long format

### dplyr

- group_by

- mutate

- join, inner, outer, anti. vs merge. relational database.

- middleware abstraction. try with sqlite as relational db.

## Assignment

### Individual

Answer the following questions using `dplyr` and `tidyr` about GHG emissions since 1970...

```{r get-CO2, results='hide'}
library(readxl)
url = 'http://edgar.jrc.ec.europa.eu/news_docs/CO2_1970-2014_dataset_of_CO2_report_2015.xls'
xls = 'data/CO2_1970-2014_dataset_of_CO2_report_2015.xls'

if (!file.exists(xls)){
  download.file(url, xls)
}
d = read_excel(xls, skip=12)
```

sources:
- [CO2 time series per region/country \| EUROPA](http://edgar.jrc.ec.europa.eu/overview.php?v=CO2ts1990-2014&sort=des9)
- [List of countries by carbon dioxide emissions - Wikipedia, the free encyclopedia](https://en.wikipedia.org/wiki/List_of_countries_by_carbon_dioxide_emissions)

- top 3 emitters in last year?  compared to rest of world?

- cumulative emissions?

- per capita? join

- join with population over time. get at emissions per capita
- rank worst 10
- get at cumulative emissions
- join with GDP. fit model, who excessively bad/goodget?
- references relevant to UCOP21:
  - [United Nations Framework Convention on Climate Change - Wikipedia, the free encyclopedia](https://en.wikipedia.org/wiki/United_Nations_Framework_Convention_on_Climate_Change#Paris_Agreement)
  - [6 Graphs Explain the World’s Top 10 Emitters | World Resources Institute](http://www.wri.org/blog/2014/11/6-graphs-explain-world%E2%80%99s-top-10-emitters)
  - [Global Emissions | Climate Change | US EPA](http://www3.epa.gov/climatechange/ghgemissions/global.html)
- just USA


### Group


