---
title: "Group Project - CRSeafood"
author: "Juliana Herrera"
date: "January 19, 2016"
output: html_document
---

#Content
I'm very interested in the effects of markets over fisheries management. My Group Project is focused in answering the overarching question **_What are the consequences of applying seafood certification programs to 
small-scale fisheries?_**, focusing on a Small scale fishery in Costa Rica. Our specific questions are: 

  + What are the factors affecting the current supply chain?
  + How do certifications affect the profits of fishers and retailers?
  + Do they affect the ratio of domestic sales to exports?

###Costa Rican Snapper fishery as a case study
The project is being developed in the North Pacific of the country, with two communities of small scale fishers who work with bottom long lines. Since 2008 the communities have been working to achieve recognition of sustainability through the adquisition of a certification. They are currently under assessment of FairTrade. 

#Techniques
In order to understand what affects the supply chain of snapper, we had to first map how is it laid out. We did this through **_systems thinking_**. The following figure shows our preliminary results:


#Data

```{r}
summary(cars)
```

You can also embed plots, for example:

```{r, echo=FALSE}
plot(cars)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
