---
title: "extra"
author: "Ben Best"
date: "January 22, 2016"
output: html_document
---

## Extra

### Get $CO_2$ Emissions

```{r get-CO2}
library(readxl)
url = 'http://edgar.jrc.ec.europa.eu/news_docs/CO2_1970-2014_dataset_of_CO2_report_2015.xls'
xls = 'data/CO2_1970-2014_dataset_of_CO2_report_2015.xls'

if (!file.exists(xls)){
  download.file(url, xls)
}
d = read_excel(xls, skip=12)
```


[CO2 time series per region/country \| EUROPA](http://edgar.jrc.ec.europa.eu/overview.php?v=CO2ts1990-2014&sort=des9)

- join with population over time. get at emissions per capita
- rank worst 10
- get at cumulative emissions
- join with GDP. fit model, who excessively bad/goodget?
- references relevant to UCOP21:
  - [United Nations Framework Convention on Climate Change - Wikipedia, the free encyclopedia](https://en.wikipedia.org/wiki/United_Nations_Framework_Convention_on_Climate_Change#Paris_Agreement)
  - [6 Graphs Explain the World’s Top 10 Emitters | World Resources Institute](http://www.wri.org/blog/2014/11/6-graphs-explain-world%E2%80%99s-top-10-emitters)
  - [Global Emissions | Climate Change | US EPA](http://www3.epa.gov/climatechange/ghgemissions/global.html)
- just USA


nycflights13



```{r}
library(dplyr)
library(nycflights13)

# Convert data to a tbl_df so that it uses dplyr's nice print method
flights <- tbl_df(flights)
flights

# Use glimpse to few some values in each column
glimpse(flights)

# filter: inspect subsets of data
# How many flights flew to Madison in 2013?
flights %>%
  filter(dest == "MSN")

# filter: multiple conditions
# How many flights flew to Madison in first week of January?
# Comma separated conditions are combined with '&'
flights %>%
  filter(dest == "MSN", month == 1, day <= 7)

# QuestionCommas in the filter statement are implicit & (and) operators. Is there anything similar for | (or)?
# Logical or statements are supported, but there's no shorthand.
flights %>%
  filter(dest == "MSN" | dest == "ORD" | dest == "MDW")
# For more complicated checks, I would try a set operation.
flights %>%
  filter(dest %in% c("MSN", "ORD", "MDW"))


# arrange: sort columns
# Sort by which airport they departed from in NYC, then year, month, day.
flights %>%
  arrange(origin, year, month, day)

# desc: reverses sorting of a column
# Find longest delayed flights to Madison.
flights %>%
  filter(dest == "MSN") %>%
  arrange(desc(dep_delay))

# select
# Select the columns you want.
flights %>%
  select(origin, year, month, day)

# select's helpers
# select has many helper functions. See ?select.
flights %>%
  select(origin, year:day, starts_with("dep"))

# negative selecting
# We can drop columns by “negating” the name. Since helpers give us column names, we can negate them too.
flights %>%
  select(-dest, -starts_with("arr"),
         -ends_with("time"))
```

Recap: Verbs for inspecting data
- convert to a `tbl_df` - nice print method
- `glimpse` - some of each column
- `filter` - subsetting
- `arrange` - sorting (desc to reverse the sort)
- `select` - picking (and omiting) columns

```{r}
# rename
# Rename columns with rename(NewName = OldName). To keep the order correct, read/remember the renaming = as “was”.
flights %>%
  rename(y = year, m = month, d = day)

# mutate
# How much departure delay did the flight make up for in the air?
flights %>%
  mutate(
    gain = arr_delay - dep_delay,
    speed = (distance / air_time) * 60,
    gain_per_hour = gain / (air_time / 60)) %>%
  select(gain:gain_per_hour)

# group_by
# Let's compute the average delay per month of flights to Madison.
# Normally–in aggregate, by or plyr's d*ply functions–you specify the grouping as an argument to the aggregation function.
aggregate(dep_delay ~ month, flights, mean,
          subset = flights$dest == "MSN")

# group_by
# In dplyr, grouping is its own action. It is done as its own step in the pipeline. Here, we filter to the flights to Madison and group_by month.
msn_by_month <- flights %>%
  filter(dest == "MSN") %>%
  group_by(month)
msn_by_month

# summarise
# Now we use summarise to compute (several) aggregate values within each group (month). summarise returns one row per group.
msn_by_month %>%
  summarise(
    flights = n(),
    avg_delay = mean(dep_delay, na.rm = TRUE),
    n_planes = n_distinct(tailnum))

# tally
# tally is a shortcut for counting number of items per group.
# Number of flights from NYC by destination by month:
flights %>%
  group_by(dest, month) %>%
  tally

# ungroup
# Remove the grouping structure with ungroup.
msn_by_month %>% ungroup

# Summarizing undoes grouping.
# Each summarise statement peels off one layer of grouping (from the right of the list of groups).
# day gets peeled off
per_day <- flights %>%
  group_by(dest, year, month, day) %>%
  summarise(flights = n())
per_day

# Peel off month grouping
per_month <- per_day %>%
  summarise(flights = sum(flights))
per_month
```


That covers 80% of dplyr
- select
- filter
- arrange
- glimpse
- rename
- mutate
- group_by, ungroup
- summarise

Other 20%

- assembly: bind_rows, bind_cols
- column-wise operations: mutate_each, summarise_each
- join tables together: left_join, right_join, inner_join, full_join
- filtering joins: semi_join, anti_join
- do: arbitrary code on each chunk
- different types of tabular data (databases, data.tables)